---
title: "Introduction"
format: html
---

## Stan

Stan is a platform for Bayesian modeling of data. It is a platform because it combines a language with a compiler and sampler, and an API that enables extensions in other statistical programming software. Today, obviously, we will talk about Stan in R. Working in R also provides an ecosystem of related packages that make preparing and interpreting your model easier.

## Installing Stan
There is more than one way to do it! I recommend installing the `cmdstanr` package.

### Requirements
You need a C++ compiler toolchain in order to use Stan through `cmdstanr` (other packages may install their own compiler[?]). If you don't have a C++ compiler toolchain, here's how to install one:

 - *Windows* - [Install Rtools: https://cran.r-project.org/bin/windows/Rtools/rtools43/rtools.html](https://cran.r-project.org/bin/windows/Rtools/rtools43/rtools.html)
 - *Mac* - Uses XCode command line tools. [Check the tools for R: https://mac.r-project.org/tools/](https://mac.r-project.org/tools/)
 - *Linux* - your package manager should be able to install `gcc`.
 
### Steps
You'll install the `cmdstanr` package, which depends on haing `cmdstan` installed. But don't worry - the R package provides a function that downloads and installs `cmdstan`. First install the R package:

```{r}
#| eval: false
# we recommend running this is a fresh R session or restarting your current session
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

Then use it to install `cmdstan`:
```{r}
#| eval: false
cmdstanr::install_cmdstan()
```



